Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-10 paper/paper.tex
quicksort in Haskell

\begin{minted}{haskell}
hylo :: (c -> f c) -> (a -> f a) -> c -> a
hylo c a = a . fmap (hylo c a). c
\end{minted}

@@

Suppose that \coq{F} is a container represented in Coq. One can think of these
\coq{F} as pairs containing the shapes and families of positions (which are, in
turn, defined in terms of decidable validity predicates).
An extension of a container \coq{F}, \coq{App F} in Coq, is then defined as
follows:
\begin{coqcode}
Record App F (X : Type)
   := { shape : Shape F; cont : {p | valid shape p} -> X }.
\end{coqcode}

@@

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App F A -> B} (or \coq{B -> App F A}).
%
Since these types now come with their own equivalence, any function that
manipulates them needs to be \emph{respectful}; i.e.\ given
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{coqcode}
forall (x y : X), R x y -> R' (f x) (f y)
\end{coqcode}

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of
\emph{setoids}, types with an associated equivalence relation, and considering
only functions that respect the associated equivalences, or \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation. We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
relation of a setoid. By default, we associate every Coq type with the standard
propositional equality, unless a different equivalence is specified (we allow
overlapping instances, and Coq's propositional equality takes the lowest
priority).
%\begin{coqcode}
%Reserved Notation "f =e g" (at level 70, no associativity).
%Class setoid A : Type :=
%  MkEquiv
%    { eqRel : A -> A -> Prop;
%      e_refl : forall x, eqRel x x;
%      e_sym : forall x y, eqRel x y -> eqRel y x;
%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
%    }.
%Notation "f =e g" := (eqRel f g).
%\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represent \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Record morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
We rely on Coq's type class mechanism to fill in the necessary equivalence
relations. Coq's code extraction mechanism will erase any occurrence
of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Inductive AppR F `{setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since this leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

@@

We prove the isomorphisms of polynomial types and the equivalent 
container extensions. As an example, we will consider the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl : Pos l -> Pos (l * r)} (resp. \coq{p_inr}) that
act as \coq{inl} (resp.\ \coq{inr}) on product positions, and
\coq{case_pos : (Pos l -> X) -> (Pos r -> X) -> Pos (l * r) -> X} that pattern
matches on the product positions, the functions that witness the isomorphism
are:
\begin{coqcode}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (p_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (p_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := case_pos (cont (fst x)) (cont (snd x)) |}.
\end{coqcode}

\noindent
Proving that the composition of these functions is the identity is
straightforward using the fact that \coq{Prod} containers only have two valid
positions.

@@

\noindent
For simplicity, we will abuse terminology, and call these
\coq{F}-algebras, instead of `\coq{App F}'-algebras.

The initial \coq{F}-algebra can be defined as follows, thanks to the use of
containers:
\begin{coqcode}
Inductive LFix F : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}

\noindent
where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:rec-schemes}).
As an example, the initial \coq{F}-algebra for the
container that is isomorpic to the functor
\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

We define \coq{LFix} as a setoid, where its equivalence relation can be
described as the least fixed point of \coq{AppR} (see
Section~\ref{sec:types-morphisms}). We define smart constructors for the
isomorphism of least fixed points as respectful morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F         l_out : LFix F ~> App F (LFix F)
\end{coqcode}

Catamorphisms are constructed, as expected, so that they structurally
deconstruct the datatype, call themselves recursively, and then compose the
result using an \coq{F}-algebra.
\begin{coqcode}
Definition cata_f (alg : Alg F A) : LFix -> A
  := fix f (x : LFix)
     := match x with
        | LFix_in ax =>
             alg {shape := shape ax; cont := fun e => f (cont ax e)}
        end.
\end{coqcode}

It is easy to show that this function is a respectful morphism of
\coq{F}-algebras. In fact, it is possible to define it as a map of the
following type:
\begin{coqcode}
  cata : forall `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}

\noindent
We prove that catamorphisms satisfy the universality property we explained in
Section~\ref{sec:rec-schemes}:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}

\noindent
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.

@@

\noindent
where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
defined as a setoid, with an equivalence relation that is the greatest fixpoint
of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
greatest fixed points:
\begin{coqcode}
g_in : App F (GFix F) ~> GFix F          g_out : GFix F ~> App F (GFix F)
\end{coqcode}

@@

Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*...*)
\end{coqcode}

\noindent
From this definition the universality property falls out:
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In words, for any \coq{F}-coalgebra, if there is any other function \coq{f}
that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the
same coalgebra.

\subsection{Mechanising Hylomorphisms}
\label{sec:rec-coalgebras}
Recall that hylomorphisms capture the concept of \emph{divide-and-conquer}
algorithms where the input is first destructured (\emph{divide}) in smaller
parts by means of a coalgebra which are computed recursively and then composed
back together (\emph{conquer}) by means of an algebra.

As we mentioned in Section~\ref{sec:rec-schemes}, given an $F$-algebra and
$F$-coalgebra, the hylomorphism is the unique solution (when it exists) to the
equation
\[
  f = a \circ F\; f \circ c
\]
As we stated earlier, a solution to this equation does not exist for an
arbitrary  algebra/coalgebra pair and, in fact, this definition cannot be
accepted by Coq.

@@

\noindent
For convenience, we package recursive coalgebras with their proofs that they
terminate for all inputs:
\begin{coqcode}
Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
\end{coqcode}
Recursive hylomorphisms are implemented in Coq recursively on the structure of
the proof for the type (\coq{RecF}) as follows:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
  : forall (x : A), RecF c x -> B
  := fix f x H
     := match c x as cx
          return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B
        with
        | cx => fun H =>
            a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
        end (RecF_inv H).
\end{coqcode}

@@

>>>>

\begin{minted}{haskell}
hylo :: Functor f => (c -> f c) -> (f a -> a) -> c -> a
hylo c a = a . fmap (hylo c a) . c
\end{minted}

 As an example, we the quicksort program can be written as an hylomorphism

% quicksort :: [Int] -> [Int]
% quicksort (x:xs) = let (xs, x, ys) = partition x xs
%                in (quicksort xs) ++ [x] ++ (quicksort ys)
%   where
%     partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r > x])

\begin{minted}{haskell}
data TreeF a k = Leaf | Branch k a k deriving Functor
\end{minted}

\begin{minted}{haskell}
divide :: [Int] -> TreeF Int [Int]
divide [] = Leaf
divide (p:xs) = Branch [l | l <- xs, l < p] p [r | r <- xs, r > p]
\end{minted}

\begin{minted}{haskell}
conquer :: TreeF Int [Int] -> [Int]
conquer Leaf = []
conquer (Branch ls x rs) = ls ++ [x] ++ rs
\end{minted}
\begin{minted}{haskell}
quicksort' = hylo divide conquer
\end{minted}

@@

An extension of a container \coq{F}, \coq{App F} in Coq, is then defined as
follows:
\begin{coqcode}
Record App F (X : Type)
   := { shape : Shape F; cont : {p | valid shape p} -> X }.
% \end{coqcode}

@@

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App A -> B} (or \coq{B -> App A}).
%
Since these types now come with their own equivalence, any function that
manipulates them needs to be \emph{respectful}; i.e.\ given
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{coqcode}
forall (x y : X), R x y -> R' (f x) (f y)
\end{coqcode}

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of
\emph{setoids}, types with an associated equivalence relation, and considering
only functions that respect the associated equivalences, or \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation. We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
relation of a setoid. By default, we associate every Coq type with the standard
propositional equality, unless a different equivalence is specified (we allow
overlapping instances, and Coq's propositional equality takes the lowest
priority).
%\begin{coqcode}
%Reserved Notation "f =e g" (at level 70, no associativity).
%Class setoid A : Type :=
%  MkEquiv
%    { eqRel : A -> A -> Prop;
%      e_refl : forall x, eqRel x x;
%      e_sym : forall x y, eqRel x y -> eqRel y x;
%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
%    }.
%Notation "f =e g" := (eqRel f g).
%\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represent \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Record morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
We rely on Coq's type class mechanism to fill in the necessary equivalence
relations. Coq's code extraction mechanism will erase any occurrence
of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Inductive AppR F `{setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

@@

We prove the isomorphisms of polynomial types and the equivalent 
container extensions. As an example, we will consider the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl : Pos l -> Pos (l * r)} (resp. \coq{p_inr}) that
act as \coq{inl} (resp.\ \coq{inr}) on product positions, and
\coq{case_pos : (Pos l -> X) -> (Pos r -> X) -> Pos (l * r) -> X} that pattern
matches on the product positions, the functions that witness the isomorphism
are:
\begin{coqcode}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (p_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (p_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := case_pos (cont (fst x)) (cont (snd x)) |}.
\end{coqcode}
Proving that the composition of these functions is the identity is
straightforward using the fact that \coq{Prod} containers only have two valid
positions.

@@

\noindent
For simplicity, we will abuse terminology, and we calling these
\coq{F}-algebras, instead of `\coq{App F}'-algebras.

The initial \coq{F}-algebra can be defined as follows, thanks to the use of
containers:
\begin{coqcode}
Inductive LFix F : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:rec-schemes}).
As an example, the initial \coq{F}-algebra for the
functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

We define \coq{LFix} as a setoid, where its equivalence relation can be
described as the least fixed point of \coq{AppR}. We define smart
constructors for the isomorphism of least fixed points as respectful morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F         l_out : LFix F ~> App F (LFix F)
\end{coqcode}

Catamorphisms are constructed, as expected, so that they structurally
deconstruct the datatype, call themselves recursively, and then compose the
result using an \coq{F}-algebra.
\begin{coqcode}
Definition cata_f F (alg : Alg F A) : LFix -> A
  := fix f (x : LFix)
     := match x with
        | LFix_in ax =>
             alg {shape := shape ax; cont := fun e => f (cont ax e)}
        end.
\end{coqcode}
It is easy to show that this function is a respectful morphism of
\coq{F}-algebras. In fact, it is possible to define it as a map of the
following type:
\begin{coqcode}
  cata : forall F `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}
We prove that catamorphisms satisfy the universality property we explained previously:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.

@@

\noindent
where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
defined as a setoid, with an equivalence relation that is the greatest fixpoint
of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
greatest fixed points
\begin{coqcode}
g_in : App F (GFix F) ~> GFix F          g_out : GFix F ~> App F (GFix F)
\end{coqcode}

@@

Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*...*)
\end{coqcode}
From this definition the universality property falls out:
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In words, for any \coq{F}-coalgebra, if there is any other function \coq{f}
that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the
same coalgebra.

\subsection{Recursive Coalgebras and Hylomorphism}
\label{sec:rec-coalgebras}
Recall that hylomorphisms capture the concept of \emph{divide-and-conquer}
algorithms where the input is first destructured (\emph{divide}) in smaller
parts by means of a coalgebra which are computed recursively and then composed
back together (\emph{conquer}) by means of an algebra.

In general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
unique solution (when it exists) to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}
As we stated earlier, a solution to this equation does not exist for an
arbitrary  algebra/coalgebra pair and, in fact, this definition cannot be
accepted by Coq.

@@

\noindent
A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra
\coq{c} such that \coq{forall x, RecF c x}.
Recursive hylomorphisms are implemented in Coq recursively on the structure of
the proof for the type (\coq{RecF}) as follows:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
  : forall (x : A), RecF c x -> B
  := fix f x H
     := match c x as cx
          return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B
        with
        | MkApp sx kx => fun H => a (MkApp sx (fun e => f (kx e) (H e)))
        end (RecF_inv H).
\end{coqcode}

@@

For convenience, we package coalgebras with their proofs that they terminate for
all inputs:
\begin{coqcode}
Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
\end{coqcode}
